Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-09-16 paper/main.tex
math aug paper

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.
\begin{definition}
  An \textcolor{action}{\textbf{action}} $act: G \times X \rightarrow X$ is a function that   satisfies the axioms of identity $act(e, x)$ for $e \in G$ and associativity $act(g, act(f, x)) = act(f \circ g, x)$ for $f,g \in G$ \cite{nlab:action}.
\end{definition}
Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, 
 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings.

@@

Sheaves are built out of presheaves, which are functors, meaning they are functions from objects of one category to objects of another category\cite{bradleyWhatFunctorDefinitions} that preserve the morphisms in each category. 
\begin{definition} A \textbf{functor} is a map $F: \mathcal{C} \rightarrow \mathcal{D}$, which means it is a function between objects $F: \textbf{ob}(\mathcal{C}) \mapsto \textbf{ob}(\mathcal{D})$ and that for every morphism $f \in Hom(C_1, C_2)$  there is a corresponding function $F: Hom(C1, C2) \mapsto Hom(F(C_1), F( C_2))$.
A \textbf{functor} must satisfy the properties 
\begin{itemize}
  \item \textit{identity}: $F(id_{C}(C)) = id_{D}(F(C))$
  \item \textit{composition}: $F(g)\circ F(f) = F(g\circ f)$ for any composable morphisms $C_{1}\xrightarrow{f} C_2$, $C_2 \xrightarrow{g} C_3$
\end{itemize}
$F(C) \in \textbf{ob}(\mathcal{D})$ denotes the object to which an object $C$ is mapped, and $F(f) \in Hom(F_(C_1), F_(C_2))$ denotes the morphism that $f$ is mapped to. 
\end{definition}

Functors preserve the domains, codomains, composition, and identities\cite{riehlCategoryTheoryContext}. Specifically, functors allow us to express which structure must be preserved when an object is transformed from one category to another. In this work, we use a presheaf functor to encode the constraints that ensure that the topological structure has been reserved.

\begin{definition}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{nlab:presheaf}. 
  \end{definition}
Specifically, we use a presheaf to describe the expectation that any set of sections defined over a larger space $\dsection \in \Gamma(\opensetc_2, \dtotal\restriction_{\openset_{2}})$ is also defined over subspaces of that space $\dsection\restriction_{\opensetc_{1}}$. We do this by defining the presheaf as the association between an openset and its corresponding set of sections
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}

The presheaf preserves continuity by stating that given a morphism $\iota$ between two objects of $\mathcal{\dbase}$ 
\begin{equation*}
  \label{eq:atct:sheaf:inclusion}
  \iota: \openset_1 \rightarrow \openset_2
\end{equation*}
there must be a corresponding restriction morphiism $\iota^*$ between the sets of sections 
\begin{equation*}
  \label{eq:atct:sheaf:restriction}
  \iota^*: \cgamma{\openset_2}{\dtotal_{\openset_2}} \rightarrow \cgamma{\openset_1}{\dtotal_{\openset_1}}
\end{equation*}
which are the outputs of the sheaf $\sheaf(\openset) = \cgamma{\openset}{\dtotal\restriction_{\openset}}$. The presheaf is contravariant because the morphisms between the input openset objects go in the opposite direction from the morphisms between the output set objects. The expectation that data containers implement inclusion and restriction morphisms means that, for example, a data container mist ensure that sections can be interpolated over any subset that the container can provide.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | c c c|}
    \hline
    &\textcolor{base}{point}/\textcolor{base}{openset}/\textcolor{base}{base space} & \textcolor{fiber}{fiber space} & \textcolor{total}{total space}\\
     &  location/subset/topology & fields &  dataset type\\
    \hline
   Data & $\dbasepoint \in \opensetc \subseteq \dbasec$ & \dfiberc & \dtotalc\\
   Visual & $\dbasepoint \in \opensetc \subseteq \dbasec$  & \vfiberc & \vtotalc\\
   Graphic & $\gbasepoint \in \opensetgc \subseteq \gbasec$ & \gfiberc & \gtotalc\\
   \hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles}}
  \label{tab:appendix:summary:objects}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l l | }
    \hline
     & \textcolor{section}{section} & \textcolor{sheaf}{sheaf} \\
     & location to record &  subset to set of possible records  \\
     \hline
  Data & $ \cgamma{\dbasec}{\dtotalc} \ni \dsectionc: \dbase \textcolor{section}{\rightarrow} \dfiberc$ & $\sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}$\\
  Visual &  $\cgamma{\dbasec}{\vtotalc} \ni \vsectionc: \dbase \textcolor{section}{\rightarrow} \vfiberc$ & $\sheafc_{\dbasec, \vtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}}$\\
  Graphic &    $\cgamma{\gbasec}{\gtotalc} \ni \gsectionc: \gbase \textcolor{section}{\rightarrow} \gfiberc$ &  $\sheafc_{\gbasec, \gtotalc}: \opensetgc \rightarrow \cgamma{\opensetc}{\gtotalc\restriction_{\opensetgc}}$ \\ 
  \hline
  \end{tabular}
  \caption{Functions that associate topological subspaces with records, discussed in \autoref{sec:atct:fb:sections} and \autoref{sec:atct:sheaves}}
  \label{tab:appendix:summary:datafunctions}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
data for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data base spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.2}
\begin{tabular}{}
  \note{changing locations, moving records to new locations, changing records}

\end{tabular}

\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
\label{tab:appendix:summary:datamod}
}
\end{table}

>>>>

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e \in G$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.
\begin{definition}
  An \textcolor{action}{\textbf{action}} \textit{act} of $G$ on $X$ is a function  $act: G \times X \rightarrow X$. An action has the properties of identity $act(e, x) = x$ for all $ x \in X$ and associativity $act(g, act(f, x)) = act(f \circ g, x)$ for $f,g \in G$ \cite{nlab:action, wki}.
\end{definition}
In \autoref{sec:artist:equivariant}, actions on data and compatible actions on graphics define structure in terms of testing whether a visualization is equivariant. Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$

 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings.

@@

Sheaves are built out of presheaves, which are functors, meaning they are functions from objects of one category to objects of another category\cite{bradleyWhatFunctorDefinitions} that preserve the morphisms in each category. 
\begin{definition} A \textbf{functor} is a map $F: \mathcal{C} \rightarrow \mathcal{D}$, which means it is a function between objects $F: \textbf{ob}(\mathcal{C}) \mapsto \textbf{ob}(\mathcal{D})$ and that for every morphism $f \in Hom(C_1, C_2)$  there is a corresponding function $F: Hom(C1, C2) \mapsto Hom(F(C_1), F( C_2))$.
A \textbf{functor} must satisfy the properties 
\begin{itemize}
  \item \textit{identity}: $F(id_{C}(C)) = id_{D}(F(C))$
  \item \textit{composition}: $F(g)\circ F(f) = F(g\circ f)$ for any composable morphisms $C_{1}\xrightarrow{f} C_2$, $C_2 \xrightarrow{g} C_3$
\end{itemize}
$F(C) \in \textbf{ob}(\mathcal{D})$ denotes the object to which an object $C$ is mapped, and $F(f) \in Hom(F_(C_1), F_(C_2))$ denotes the morphism that $f$ is mapped to. 
\end{definition}
Functors preserve the domains, codomains, composition, and identities\cite{riehlCategoryTheoryContext}.  
\begin{definition}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{nlab:presheaf}. 
\end{definition}
Contravariance means that the morphisms between the input openset objects go in the opposite direction from the morphisms between the output set objects. We specify behavior using functors because it allows us to express which structure must be preserved when an object is transformed from one category to another.

We use a presheaf functor to encode the binding between an openset and corresponding sets of sections 
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
and the morphisms preserved by the functor describe how subsets of the data are expected to behave. The presheaf is contravariant because the inclusion morphisms between input objects 
\begin{equation*}
  \label{eq:atct:sheaf:inclusion}
  \iota: \opensetc_1 \rightarrow \opensetc_2
\end{equation*}
are defined such that they correspond to the partial ordering $\openset_1 \subseteq \openset_2$, but the restriction morphisms $\iota^*$ between the sets of sections 
\begin{equation*}
  \label{eq:atct:sheaf:restriction}
  \iota^*: \cgamma{\opensetc_2}{\dtotalc_{\opensetc_2}} \rightarrow \cgamma{\opensetc_1}{\dtotalc_{\openset_c1}}
\end{equation*}
go in the opposite direction because any arbitrary function defined on an openset $\dsection\restriction_{\openset_2}$ must be definable on a subset of that space $\dsection\restriction_{\openset_1}$. The expectation that data containers implement inclusion and restriction morphisms means that, for example, a data container must ensure that continuous sections can be interpolated.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | c c c|}
    \hline
    &\textcolor{base}{point}/\textcolor{base}{openset}/\textcolor{base}{base space} & \textcolor{fiber}{fiber space} & \textcolor{total}{total space}\\
     &  location/subset/topology & fields &  dataset type\\
    \hline
   Data & $\dbasepointc \in \opensetc \subseteq \dbasec$ & \dfiberc & \dtotalc\\
   Visual & $\dbasepointc \in \opensetc \subseteq \dbasec$  & \vfiberc & \vtotalc\\
   Graphic & $\gbasepoint \in \opensetgc \subseteq \gbasec$ & \gfiberc & \gtotalc\\
   \hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles}}
  \label{tab:appendix:summary:objects}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l l | }
    \hline
     & \textcolor{section}{section} & \textcolor{sheaf}{sheaf} \\
     & record at location & set of possible records for subset \\
     \hline
  Data & $ \cgamma{\dbasec}{\dtotalc} \ni \dsectionc: \dbase \textcolor{section}{\rightarrow} \dfiberc$ & $\sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}$\\
  Visual &  $\cgamma{\dbasec}{\vtotalc} \ni \vsectionc: \dbase \textcolor{section}{\rightarrow} \vfiberc$ & $\sheafc_{\dbasec, \vtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}}$\\
  Graphic &    $\cgamma{\gbasec}{\gtotalc} \ni \gsectionc: \gbase \textcolor{section}{\rightarrow} \gfiberc$ &  $\sheafc_{\gbasec, \gtotalc}: \opensetgc \rightarrow \cgamma{\opensetc}{\gtotalc\restriction_{\opensetgc}}$ \\ 
  \hline
  \end{tabular}
  \caption{Functions that associate topological subspaces with records, discussed in \autoref{sec:atct:fb:sections} and \autoref{sec:atct:sheaves}}
  \label{tab:appendix:summary:datafunctions}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
 record for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)$  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data base spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|}

\end{tabular}
\note{changing locations, moving records to new locations, changing records}
\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
\label{tab:appendix:summary:datamod}
}
\end{table}
